---
title: "penguin_HMM_modeling"
author: "Taylor Azizeh"
date: "2025-02-28"
output: html_document
---

**Title**: Hidden Markov Modelling for foraging emperor penguins   
**Contact**: taylorazizeh@gmail.com  
**Author**: Taylor Azizeh (adapted from code by Parker Forman)
**Description**: This code walks through fitting a Hidden Markov model to foraging emperor penguin data. This code runs through: importing and reformatting data, visualizing the data,


STEP I: Load and reformat data.
```{r}
# Load libraries
library(data.table)
library(dplyr)
library(lubridate)
library(momentuHMM)
library(terra) # for plotting
library(sf) # for plotting
library(MASS)  # For fitting negative binomial models
library(performance)
library(ggplot2)
library(ggspatial) # for plotting
library(viridis) # for plotting
library(geosphere)
library(readr)      # if you are using read_csv/write_csv
library(beepr)      # if you want the beep()
```

```{r}
# Because your environment lacks logLik.momentuHMM, define it:
logLik.momentuHMM <- function(object, ...) {
  if(!inherits(object, "momentuHMM"))
    stop("Need a momentuHMM object")
  # momentuHMM typically stores negative log-likelihood in object$mod$minimum
  -object$mod$minimum
}

# Similarly, define a BIC method for momentuHMM:
BIC.momentuHMM <- function(object, ...) {
  if(!inherits(object, "momentuHMM"))
    stop("Need a momentuHMM object")
  # SAPple size
  n <- nrow(object$data)
  # Number of parameters (estimated)
  k <- length(object$mod$estimate)
  # Log-likelihood:
  ll <- logLik.momentuHMM(object)
  # BIC calculation:
  -2 * ll + k * log(n)
}
```

```{r}
# Import GPS data
gps <- read.csv('/Users/taylorazizeh/Documents/research/active/Emperor penguins/data/cleaned/gps interpolated/fixed timestamps/cleaned_GPS_data_all_years.csv')
head(gps)

# Import prey capture data
prey <- read.csv('/Users/taylorazizeh/Documents/research/active/emperor penguins/results/MATLAB/final/prey_capture_results.csv')
head(prey)

# Subset prey data
myvars <- c("BirdID", "DiveNumber", "DiveStartTime", "DiveDuration", "TotalEvents", "ODBA_Total", "DiveType")
prey <- prey[myvars]

# Remove "x" in front of IDs from MATLAB data
prey <- prey %>%
  mutate(BirdID = sub("^x", "", BirdID))

# Remove no GPS birds from dataset
prey <- prey %>%
  filter(!(BirdID %in% c("22EP_310c", "22EP_308c")))

# Set IDs
penguinIDs <- unique(prey$BirdID)
print(penguinIDs)
```

``` {r}
# Rename and format
gps$Timestamp <- as.POSIXct(gps$Timestamp)
prey$DiveStartTime <- as.POSIXct(prey$DiveStartTime, format = "%d-%b-%Y %H:%M:%OS")

# Check class
str(gps$Timestamp)
str(prey$DiveStartTime)

# Rename DiveStartTime to Timestamp
prey <- prey %>% 
  rename(Timestamp = DiveStartTime)
```

```{r}
# Set as data tables
prey <- as.data.table(prey)
gps <- as.data.table(gps)

# Pair GPS and prey data
setkey(gps, ID, Timestamp)
setkey(prey, BirdID, Timestamp)

# Perform a rolling join within each bird ID:
paired_data <- gps[prey, roll = "nearest"] # For each dive, the join finds the nearest GPS point in time

# Display the paired data
print(paired_data)
```

STEP II: Visualize the data.
```{r}
# Dive duration (should be between 20s and 1000s)
hist(paired_data$DiveDuration)
summary(paired_data$DiveDuration)

# Total prey capture events
hist(paired_data$TotalEvents)
summary(paired_data$TotalEvents)

# Total ODBA
hist(paired_data$ODBA_Total)
summary(paired_data$ODBA_Total)

# There is one outrageous outlier so let's check that out and remove it
max(paired_data$ODBA_Total) # This is obviously erroenous
paired_data <- paired_data %>%
  slice(-14628)  # Removes row with outlier

# Total prey capture events
hist(paired_data$TotalEvents)
summary(paired_data$TotalEvents)

# Total ODBA
hist(paired_data$ODBA_Total)
summary(paired_data$ODBA_Total)

# That looks much better!
```

```{r}
# Normalize prey capture events relative to each penguin's track
paired_data <- paired_data %>%
  group_by(ID) %>%
  mutate(
    ScaledTotalEvents = scales::rescale(TotalEvents, to = c(0, 1)) # Scale 0 to 1 for each penguin
  ) %>%
  ungroup()

# Load IBCSO bathymetry raster
ibcso_bedrock_path <- "/Users/taylorazizeh/Documents/research/active/Emperor penguins/data/cleaned/bathymetry/IBCSO_v2_bed_RGB.tif" 
ibcso_ice_path <- "/Users/taylorazizeh/Documents/research/active/Emperor penguins/data/cleaned/bathymetry/IBCSO_v2_ice-surface_RGB.tif"

# Convert to a raster
ibcso_bedrock <- rast(ibcso_bedrock_path)
ibcso_ice <- rast(ibcso_ice_path)

# Define region of interest
xmin <- -100000
xmax <- 500000
ymin <- -1550000
ymax <- -1250000
extent_of_interest <- ext(xmin, xmax, ymin, ymax)

# Crop bathymetry and ice rasters
ibcso_bedrock_cropped <- crop(ibcso_bedrock, extent_of_interest)
ibcso_ice_cropped <- crop(ibcso_ice, extent_of_interest)

# Convert bathymetry raster to data frame for ggplot
bedrock_df <- as.data.frame(ibcso_bedrock_cropped, xy = TRUE) %>%
  mutate(color = rgb(
    IBCSO_v2_bed_RGB_1 / 255,
    IBCSO_v2_bed_RGB_2 / 255,
    IBCSO_v2_bed_RGB_3 / 255
  ))

# Convert ice raster to data frame for ggplot
ice_df <- as.data.frame(ibcso_ice_cropped, xy = TRUE) %>%
  mutate(color = rgb(
    `IBCSO_v2_ice-surface_RGB_1` / 255,
    `IBCSO_v2_ice-surface_RGB_2` / 255,
    `IBCSO_v2_ice-surface_RGB_3` / 255
  ))
```

```{r}
# Convert penguin data to sf object (WGS84)
pen_data_sf <- st_as_sf(
  paired_data,
  coords = c("Lon", "Lat"),
  crs = 4326
)

# Transform penguin data to match IBCSO CRS
pen_data_sf <- st_transform(pen_data_sf, crs(ibcso_bedrock_cropped))

# Convert back to data frame for ggplot
pen_data_transformed <- pen_data_sf %>%
  st_coordinates() %>%
  as.data.frame() %>%
  bind_cols(pen_data_sf %>% st_drop_geometry()) %>%
  rename(x = X, y = Y)

# Plotting
fig2 <- ggplot() +
  # Add bathymetry raster
  geom_raster(
    data = bedrock_df,
    aes(x = -x, y = -y, fill = color),
    show.legend = FALSE
  ) +
  scale_fill_identity() +  # Use the 'color' column directly
  
  # Add ice raster with transparency
  geom_raster(
    data = ice_df,
    aes(x = -x, y = -y, fill = color),
    alpha = 0.7,
    show.legend = FALSE
  ) +
  
  # Add penguin tracks, color-coded by ID
  geom_point(
    data = pen_data_transformed,
    aes(x = -x, y = -y, color = ScaledTotalEvents),
    size = 0.8,
    alpha = 0.8
  ) +
    
  # Color scale for relative prey capture events
  scale_color_viridis_c(name = "Rel. Prey Capture") +
  
  # Add scale bar and north arrow
  annotation_scale(location = "bl", width_hint = 0.2) +
  annotation_north_arrow(location = "tr",
                         style = north_arrow_orienteering(),
                         height = unit(1.0, "cm"),
                         width = unit(1.0, "cm"),
                         pad_x = unit(0.25, "cm"),
                         pad_y = unit(0.25, "cm")) +
  
  # Add titles and labels
  labs(
    title = "Penguin Tracks with Relative Prey Capture ",
    x = "Projected X (meters)",
    y = "Projected Y (meters)"
  ) +
  
  # Coordinate system and theme
  coord_sf(crs = st_crs(ibcso_bedrock_cropped), expand = FALSE) +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 8),
        legend.key.width = unit(0.05, "lines"),
        axis.text = element_text(size = 6),
        axis.title = element_text(size = 8)) +
  guides(color = guide_legend(ncol = 1))  # Adjust columns

print(fig2)

# Save Figure 2
ggsave(fig2, 
       filename = "figure2.jpeg",
       device = "jpeg",
       path = "/Users/taylorazizeh/Documents/research/active/Emperor penguins/figures/figures",
       height = 5, width = 7, units = "in")
```

STEP III: Test for overdispersion and run HMM.
```{r}
# Test for overdispersion
test_nb <- glm.nb(TotalEvents ~ 1, data = paired_data)
check_overdispersion(test_nb)
```

```{r}
# Prepare data for HMM
df <- paired_data %>%
  dplyr::select(ID, Lon, Lat, Timestamp, TotalEvents, ODBA_Total, DiveType)

df <- as.data.frame(df)

# Check it
head(df)

# Use the function prepData
df <- prepData(df,
               type = c("LL"),
               coordNames = c("Lon", "Lat"),
               covNames = c("TotalEvents", "ODBA_Total", "DiveType"))
# Check it
head(df)
plot(df, compact = TRUE, ask = FALSE)
```

```{r}
# Make histogram of step lengths and angles
hist(df$step,
     xlab = "step length",
     main = "")
summary(df$step) # Maximum is 35

hist(df$angle,
     xlab = "turning angle",
     main = "")
summary(df$angle)
```

```{r}
# Define output folder
dir_model_out <- "/Users/taylorazizeh/Documents/research/active/Emperor penguins/results/HMM"

# We'll define custom state names = c("Resting","Foraging","Transit")
fitModel_SA <- function(df) {
  nbStates <- 3
  dist_list <- list(step="gamma", angle="vm")

  step_mean <- mean(df$step, na.rm=TRUE)
  step_sd   <- sd(df$step, na.rm=TRUE)
  stepPar0  <- c(step_mean*0.5, step_mean, step_mean*1.5,
                 step_sd*0.5,   step_sd,   step_sd*1.5)

  anglePar0 <- c(rep(0, nbStates), rep(1, nbStates))

  Par0 <- list(step=stepPar0, angle=anglePar0)

  fit <- tryCatch({
    momentuHMM:::fitHMM.momentuHMMData(
      data         = df,
      nbStates     = nbStates,
      dist         = dist_list,
      Par0         = Par0,
      stateNames   = c("Resting","Foraging","Transit"),
      estAngleMean = list(angle=TRUE),
      formula      = ~1
    )
  }, error=function(e){
    message("Error in fitModel_SA: ", e$message)
    NULL
  })
  fit
}

#################################  Helper: assign biological state labels to states 1, 2, 3
assignStateLabels <- function(fit) {
  step_means  <- fit$mle$step[1, ]
  angle_kappa <- fit$mle$angle[2, ]

  label_vector <- rep(NA, 3)

  # Transit: highest step length and high directional persistence
  transit_idx <- which.max(step_means)
  label_vector[transit_idx] <- "Transit"

  # Among remaining, Resting = shortest steps + most directed (highest kappa)
  remaining <- setdiff(1:3, transit_idx)
  kappa_sub <- angle_kappa[remaining]
  step_sub  <- step_means[remaining]

  resting_idx <- remaining[which.max(kappa_sub / step_sub)]  # smallest step + highest directionality
  label_vector[resting_idx] <- "Resting"

  # Remaining = Foraging (short step, high turning = lowest kappa)
  foraging_idx <- setdiff(1:3, c(transit_idx, resting_idx))
  label_vector[foraging_idx] <- "Foraging"

  return(label_vector)
}

################################# Fit Step-Angle-Prey
# Define SAP model function
fitModel_SAP <- function(df) {
  nbStates <- 3  # Three states: Resting, Foraging, Transit

  dist_list <- list(step="gamma", angle="vm", TotalEvents="pois") 

  step_mean <- mean(df$step, na.rm=TRUE)
  step_sd   <- sd(df$step, na.rm=TRUE)
  stepPar0  <- c(step_mean*0.5, step_mean, step_mean*1.5,
                 step_sd*0.5,   step_sd,   step_sd*1.5)

  anglePar0 <- c(rep(0, nbStates), rep(1, nbStates))
  
  prey_mean <- mean(df$TotalEvents, na.rm=TRUE)
  prey_sd   <- sd(df$TotalEvents, na.rm=TRUE)
  preyPar0  <- c(prey_mean*0.5, prey_mean, prey_mean*1.5)

  Par0 <- list(step=stepPar0, angle=anglePar0, TotalEvents=preyPar0)

  fit <- tryCatch({
    momentuHMM:::fitHMM.momentuHMMData(
      data         = df,
      nbStates     = nbStates,
      dist         = dist_list,
      Par0         = Par0,
      stateNames   = c("Resting","Foraging","Transit"),
      estAngleMean = list(angle=TRUE),
      formula      = ~1
    )
  }, error=function(e){
    message("Error in fitModel_SAP: ", e$message)
    NULL
  })
  fit
}

################################# Fit Step-Angle-Prey (ZI)
fitModel_SAP_ZI <- function(hmm_data) {
  nbStates <- 3  # Three states: Resting, Foraging, Transit

  # Skip if no variation in prey capture events
  if (sd(hmm_data$TotalEvents, na.rm=TRUE) == 0) {
    message("No variation in TotalEvents => skip SAP-ZI.")
    return(NULL)
  }

  # Use Negative Binomial (handles zero-inflation naturally)
  dist_list <- list(
    step = "gamma",
    angle = "vm",
    TotalEvents = "negbinom"  # No need for explicit zero-inflation parameter
  )

  # Step parameters
  step_mean <- mean(hmm_data$step, na.rm=TRUE)
  step_sd   <- sd(hmm_data$step, na.rm=TRUE)
  stepPar0  <- c(step_mean*0.5, step_mean, step_mean*1.5,
                 step_sd*0.5,   step_sd,   step_sd*1.5)

  # Angle parameters
  anglePar0 <- c(rep(0, nbStates), rep(1, nbStates))

  # Prey capture parameters (Negative Binomial)
  prey_mean <- mean(hmm_data$TotalEvents, na.rm=TRUE)
  prey_sd   <- sd(hmm_data$TotalEvents, na.rm=TRUE)

  # Negative Binomial requires: 3 mean params (mu) + 3 dispersion params (size)
  preyPar0  <- c(
    prey_mean*0.5, prey_mean, prey_mean*1.5,  # Mean prey capture rate (mu)
    2, 2, 2  # Dispersion parameters (size)
  )

  Par0 <- list(
    step = stepPar0,
    angle = anglePar0,
    TotalEvents = preyPar0
  )

  # Corrected Design Matrix (must include mu and size)
  DM_list <- list(
    TotalEvents = list(
      mu   = ~1,  # Mean prey capture rate (lambda)
      size = ~1   # Overdispersion (models excess zeros)
    )
  )

  # Fit HMM with Negative Binomial
  fit <- tryCatch({
    momentuHMM:::fitHMM.momentuHMMData(
      data       = hmm_data,
      nbStates   = nbStates,
      dist       = dist_list,
      Par0       = Par0,
      DM         = DM_list,  # Now correctly includes both mu & size
      stateNames = c("Resting","Foraging","Transit"),
      estAngleMean = list(angle=TRUE),
      formula    = ~1
    )
  }, error=function(e){
    message("Error in fitModel_SAP_ZI: ", e$message)
    NULL
  })

  fit
}


################################# Prepare & fit
prepAndFitHMM <- function(df) {
  baseObj <- df  # Use the already prepared dataset

  if (!all(c("step", "angle", "TotalEvents") %in% names(baseObj))) {
    stop("Error: The dataset must contain 'step', 'angle', and 'TotalEvents' columns.")
  }

  # Remove rows with NA in step or angle
  baseObj <- baseObj %>%
    filter(!is.na(step), !is.na(angle))

  if(nrow(baseObj) < 3) {
    return(list(SA=NULL, SAP=NULL, SAP_ZI=NULL, hmmData=baseObj))
  }

  # Fit SA & SAP models
  fitSA  <- fitModel_SA(baseObj)
  fitSAP <- fitModel_SAP(baseObj)

  # Fit SAP-ZI (zero-inflated prey capture) only if there’s variation in TotalEvents
  fitSAP_ZI <- if (sd(baseObj$TotalEvents, na.rm=TRUE) > 0) {
    fitModel_SAP_ZI(baseObj)
  } else {
    message("No variation in TotalEvents => skipping SAP-ZI.")
    NULL
  }

  list(SA=fitSA, SAP=fitSAP, SAP_ZI=fitSAP_ZI, hmmData=baseObj)
}

################################# We'll keep the existing summary of "best" models:
bestModels   <- list()
summaryTable <- data.frame(
  ID         = character(),
  Best_Model = character(),
  AIC        = numeric(),
  BIC        = numeric(),
  stringsAsFactors=FALSE
)

# Summaries for ALL SA & SAP models:
summaryAllModels <- data.frame(
  ID    = character(),
  Model = character(),
  AIC   = numeric(),
  BIC   = numeric(),
  stringsAsFactors=FALSE
)

# Load patchwork if needed
if (!requireNamespace("patchwork", quietly = TRUE)) install.packages("patchwork")
library(patchwork)

# Fit all three models, compare, and only plot the best fit
for (penID in penguinIDs) {
  cat("\n============================================\n")
  cat("PROCESSING PENGUIN:", penID, "\n")
  cat("============================================\n")

  df_filtered <- df %>% filter(ID == penID)
  if (nrow(df_filtered) == 0) {
    cat("No data for", penID, "\n")
    next
  }

  # Fit models
  fit_list   <- prepAndFitHMM(df_filtered)
  fitSA      <- fit_list$SA
  fitSAP     <- fit_list$SAP
  fitSAP_ZI  <- fit_list$SAP_ZI
  baseObj    <- fit_list$hmmData
  baseObj$Timestamp <- as.POSIXct(baseObj$Timestamp, format = "%Y-%m-%d %H:%M:%S", tz = "UTC")

  # Store valid models
  fittedModels <- list()
  if (!is.null(fitSA)     && inherits(fitSA, "momentuHMM"))     fittedModels[["SA"]]     <- fitSA
  if (!is.null(fitSAP)    && inherits(fitSAP, "momentuHMM"))    fittedModels[["SAP"]]    <- fitSAP
  if (!is.null(fitSAP_ZI) && inherits(fitSAP_ZI, "momentuHMM")) fittedModels[["SAP_ZI"]] <- fitSAP_ZI

  if (length(fittedModels) == 0) {
    cat("No valid models for", penID, "\n")
    next
  }

  # Get AIC/BIC values
  modelAICs <- sapply(fittedModels, function(m) tryCatch(AIC(m), error = function(e) NA_real_))
  modelBICs <- sapply(fittedModels, function(m) tryCatch(BIC(m), error = function(e) NA_real_))

  # Append to summary of all models
  for (mName in names(fittedModels)) {
    summaryAllModels <- rbind(summaryAllModels, data.frame(
      ID    = paste0("Pen", penID),
      Model = mName,
      AIC   = modelAICs[mName],
      BIC   = modelBICs[mName],
      stringsAsFactors = FALSE
    ))
  }

  # Identify best model by AIC
  if (all(is.na(modelAICs))) {
    cat("All model AICs NA — skipping", penID, "\n")
    next
  }

  bestModelType <- names(which.min(modelAICs))
  bestModel     <- fittedModels[[bestModelType]]
  bestAIC       <- modelAICs[bestModelType]
  bestBIC       <- modelBICs[bestModelType]

  cat("\n>>> Best model for", penID, "is", bestModelType,
      "with AIC =", bestAIC, "| BIC =", bestBIC, "\n")

  # Save best model info
  summaryTable <- rbind(summaryTable, data.frame(
    ID         = paste0("Pen", penID),
    Best_Model = bestModelType,
    AIC        = bestAIC,
    BIC        = bestBIC,
    stringsAsFactors = FALSE
  ))
  bestModels[[paste0("Pen", penID)]] <- bestModel

  # Decode and assign behavioral state labels
  decodedStates <- viterbi(bestModel)
  
  # Use helper to assign biological labels
  state_labels <- assignStateLabels(bestModel)
  
  # Validation printout
  cat("→ Step means:", paste(round(bestModel$mle$step[1, ], 2), collapse = ", "), "\n")
  cat("→ Angle kappa:", paste(round(bestModel$mle$angle[2, ], 2), collapse = ", "), "\n")
  cat("→ Assigned labels:", paste(state_labels, collapse = ", "), "\n")
  
  # Check for errors or NA in labeling
  if (length(state_labels) != 3 || any(is.na(state_labels))) {
    warning("Invalid state_labels. Falling back to generic labels.")
    state_labels <- c("State1", "State2", "State3")
  }
  
  # Assign labels to states
  decoded_named <- factor(decodedStates, levels = 1:3, labels = state_labels)
  
  # Reclassify "Resting" during foraging-type dives as "Foraging"
  if ("DiveType" %in% names(baseObj)) {
    decoded_named <- as.character(decoded_named)  # avoid factor/ifelse mismatch
  
    reclass_idx <- decoded_named == "Resting" & baseObj$DiveType %in% c("Epipelagic", "Mesopelagic", "Benthic")
    decoded_named[reclass_idx] <- "Foraging"
  
    # Report how many were reclassified
    num_reclassified <- sum(reclass_idx, na.rm = TRUE)
    cat("✔ Reclassified", num_reclassified, "Resting → Foraging (due to dive type)\n")
  
    # Finalize factor with expected levels
    decoded_named <- factor(decoded_named, levels = c("Resting", "Foraging", "Transit"))
  }
  
  # Assign to baseObj
  baseObj$DecodedState <- decoded_named
  
  # Diagnostic: check final state counts
  cat("✔ DecodedState counts:\n")
  print(table(baseObj$DecodedState, useNA = "always"))
  
  # Save decoded CSV
  decoded_csv <- file.path(dir_model_out, paste0("Pen", penID, "_DecodedStates_", bestModelType, ".csv"))
  data.table::fwrite(baseObj, decoded_csv)
  cat("✔ Saved decoded states to:", decoded_csv, "\n")

  # Spatial plot if x/y available
  if (all(c("x", "y") %in% names(baseObj))) {
    baseObj <- baseObj %>% filter(!is.na(x), !is.na(y))

    # Plot 1: Colored by HMM state
    p_state <- ggplot(baseObj, aes(x = x, y = y)) +
      geom_path(color = "grey90") +
      geom_point(aes(color = DecodedState), size = 2, alpha = 0.9) +
      scale_color_manual(
        values = c(
          "Resting" = "lightblue",
          "Foraging" = "darkblue",
          "Transit" = "lightgrey"
        ),
        name = "State"
      ) +
      labs(
        title = "Penguin Track Colored by HMM States",
        x = "Longitude", y = "Latitude"
      ) +
      coord_fixed(ratio = 1.3) +
      theme_minimal()
    
    # Plot 2: Colored by prey capture events
    p_prey <- ggplot(baseObj, aes(x = x, y = y)) +
      geom_path(color = "grey90") +
      geom_point(aes(color = TotalEvents), size = 2, alpha = 0.9) +
      scale_color_viridis_c(
        option = "plasma",
        name = "Prey Capture Events",
        direction = -1  # darkest colors = highest events
      ) +
      labs(
        title = "Penguin Track Colored by Prey Capture Events",
        x = "Longitude", y = "Latitude"
      ) +
      coord_fixed(ratio = 1.3) +
      theme_minimal()
    
    # Set consistent x/y limits
    xlim_vals <- range(baseObj$x, na.rm = TRUE)
    ylim_vals <- range(baseObj$y, na.rm = TRUE)

    p_state <- p_state + xlim(xlim_vals) + ylim(ylim_vals)
    p_prey  <- p_prey  + xlim(xlim_vals) + ylim(ylim_vals)

    # Combine and save side-by-side
    combined_map <- p_state + p_prey + patchwork::plot_layout(nrow = 2)
    map_out <- file.path(dir_model_out, paste0("Pen", penID, "_HMM_vs_Prey_Map_", bestModelType, ".png"))
    ggsave(map_out, combined_map, width = 10, height = 14, dpi = 300)
    cat("✔ Combined state/prey map saved to:", map_out, "\n")
  }
}

# Save full model summary
if (nrow(summaryAllModels) > 0) {
  allModels_csv <- file.path(dir_model_out, "All_Penguin_Model_Results.csv")
  readr::write_csv(summaryAllModels, allModels_csv)
  cat("\n✔ All model AIC/BIC results saved to:", allModels_csv, "\n")
}

# Save best models as RDS
models_rds <- file.path(dir_model_out, "Best_HMM_Models.rds")
saveRDS(bestModels, models_rds)
cat("\n✔ Best HMM models saved to:", models_rds, "\n")
```
