---
title: "Stat analysis prey capture per dive type"
output: html_document
date: "2024-08-16"
---




# Load libraries.
``` {r libraries}
library(tidyr)
library(dplyr)
library(readr)
library(broom)
library(ggplot2) # for graphing
library(FSA) # for Dunn's test
library(purrr)
library(car)
```

# Set working directory and import data.
```{r load}
# Set working directory
setwd("/Users/taylorazizeh/Documents/Research/Moss Landing Marine Laboratories/Emperor penguins/Statistics/Prey capture per dive type")

# Load the data
df_2019_0_15 <- read_csv("2019_prey_capture_0.15.csv")
df_2019_0_30 <- read_csv("2019_prey_capture_0.30.csv")
df_2022_0_15 <- read_csv("2022_prey_capture_0.15.csv")
df_2022_0_30 <- read_csv("2022_prey_capture_0.30.csv")
```

# Reformat data.
``` {r}
# Add year and threshold columns
df_2019_0_15 <- df_2019_0_15 %>% mutate(year = 2019, threshold = 0.15)
df_2019_0_30 <- df_2019_0_30 %>% mutate(year = 2019, threshold = 0.30)
df_2022_0_15 <- df_2022_0_15 %>% mutate(year = 2022, threshold = 0.15)
df_2022_0_30 <- df_2022_0_30 %>% mutate(year = 2022, threshold = 0.30)

# Combine all data into one dataframe
combined_data <- bind_rows(df_2019_0_15, df_2019_0_30, df_2022_0_15, df_2022_0_30)

# Reshape data to long format
long_data <- combined_data %>%
  pivot_longer(cols = c("Transit", "Epipelagic", "Mesopelagic", "Benthic"),
               names_to = "dive_type",
               values_to = "prey_capture")
```

# Check data for normality using Shapiro-Wilkes test.
``` {r}
# Check for normality
# Apply Shapiro-Wilk test for each combination of dive_type, year, and threshold
shapiro_results <- long_data %>%
  group_by(dive_type, year, threshold) %>%
  summarize(p_value = shapiro.test(prey_capture)$p.value)

# Reformat p-values for easier reading
shapiro_results$p_value <- format(shapiro_results$p_value, scientific = FALSE)

# View the results
print(shapiro_results)
```

# Check histograms for distribution.
``` {r}
# Create histograms for prey capture by dive_type, year, and threshold
ggplot(long_data, aes(x = prey_capture)) +
  geom_histogram(bins = 30, fill = "blue", color = "black", alpha = 0.7) +
  facet_grid(dive_type ~ year + threshold) +
  labs(title = "Histograms of Prey Capture by Dive Type, Year, and Threshold",
       x = "Prey Capture Attempts",
       y = "Frequency") +
  theme_minimal()
```

# Run a Levene's test for variances.
``` {r}
# prey_capture is the dependent variable, and DiveDurationCategory is the grouping variable
levene_test_result <- leveneTest(prey_capture ~ dive_type, data = long_data)

# View the results of Levene's Test
print(levene_test_result)
```

# Run Krusal-Wallis test.
``` {r}
# Perform Kruskal-Wallis test for each combination of year and threshold
kruskal_results <- long_data %>%
  group_by(year, threshold) %>%
  summarize(
    kruskal_test = list(kruskal.test(prey_capture ~ dive_type)),
    test_statistic = kruskal_test[[1]]$statistic,
    p_value = kruskal_test[[1]]$p.value
  )

# Reformat p-values for easier reading
kruskal_results <- kruskal_results %>%
  mutate(p_value = format(p_value, scientific = FALSE))

# View the results
print(kruskal_results)
```

# Run post-hoc pairwise comparison.
``` {r}
# Perform Dunn's test with Bonferroni adjustment for p-values
dunn_test_results <- dunnTest(prey_capture ~ dive_type, data = long_data, method = "bonferroni")

# Make p-values easier to read
dunn_test_results$P.adj <- format(dunn_test_results$P.adj, scientific = FALSE)

# View the results of Dunn's test
print(dunn_test_results)

# Extract significant comparisons
significant_results <- dunn_test_results$res %>%
  filter(P.adj < 0.05)

# Reformat p-values for easier reading
significant_results$P.adj <- format(significant_results$P.adj, scientific = FALSE)

# View significant pairwise comparisons
print(significant_results)
```


# Create meaningful figure.
``` {r}
# Extract the results from the Dunn's test object
heatmap_data <- dunn_test_results$res %>%
  separate(Comparison, into = c("Group1", "Group2"), sep = " - ") %>%
  mutate(Significance = case_when(
    P.adj < 0.001 ~ "***",
    P.adj < 0.01 ~ "**",
    P.adj < 0.05 ~ "*",
    TRUE ~ "ns"
  )) %>%
  # Join the year and threshold information back to the heatmap_data
  left_join(select(long_data, dive_type, year, threshold), by = c("Group1" = "dive_type"))


# Create the heatmap with blue gradients, faceted by year and threshold
ggplot(heatmap_data, aes(x = Group1, y = Group2, fill = Z)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Significance), color = "white", size = 5) +
  scale_fill_gradient(low = "lightblue", high = "darkblue", 
                      name = "Z-Score") +
  labs(title = "Heatmap of Pairwise Comparisons from Dunn's Test",
       x = "Dive Type 1", y = "Dive Type 2") +
  facet_grid(year ~ threshold) +  # Facet by year and threshold
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title = element_blank())
```

# Benthic vs. pelagic test.
``` {r}
# Combine epipelagic and mesopelagic
long_data_pelagic <- long_data %>%
  mutate(dive_type = ifelse(dive_type %in% c("Mesopelagic", "Epipelagic"), "Pelagic", dive_type)) %>%
  group_by(year, threshold, dive_type) %>%
  summarize(prey_capture = sum(prey_capture, na.rm = TRUE))
```

# Run Shapiro-Wilk on new group
``` {r}
# Recheck normality for Pelagic and Benthic categories
shapiro_results_pelagic_benthic <- long_data_pelagic %>%
  filter(dive_type %in% c("Pelagic", "Benthic")) %>%
  group_by(dive_type) %>%
  summarize(p_value = shapiro.test(prey_capture)$p.value)

# Reformat p-value for easier reading
shapiro_results_pelagic_benthic$p_value <- format(shapiro_results_pelagic_benthic$p_value, scientific = FALSE)

# View the Shapiro-Wilk test results
print(shapiro_results_pelagic_benthic)
```

# Check variances
``` {r}
# Perform Levene's Test to check for equal variances
filtered_data <- long_data_pelagic %>%
  filter(dive_type %in% c("Pelagic", "Benthic"))

levene_test <- leveneTest(prey_capture ~ dive_type, data = filtered_data)

# View the results of Levene's Test
print(levene_test)
```

# Since data are normally distributed and the variances are unequal, we can run a Welch's t-test
``` {r}
# Perform Welch's t-test, which does not assume equal variances
welch_t_test_pelagic_benthic <- t.test(prey_capture ~ dive_type, data = filtered_data, var.equal = FALSE)

# View the results of Welch's t-test
print(welch_t_test_pelagic_benthic)
```

# Graph results.
``` {r}
ggplot(filtered_data, aes(x = dive_type, y = prey_capture, fill = dive_type)) +
  geom_violin(trim = FALSE) +
  geom_boxplot(width = 0.1, fill = "white") +
  scale_y_log10() +  # Log transformation of the y-axis
  labs(title = "Distribution of Prey Capture Attempts by Dive Class (Log Scale)",
       x = "Dive Type", y = "Prey Capture Attempts (Log Scale)") +  # Remove legend title
  theme_minimal() +
  scale_fill_brewer(palette = "Blues") +
  theme(
    plot.title = element_text(size = 9),  # Adjust title text size
    axis.title = element_text(size = 9),  # Adjust axis labels text size
    axis.text = element_text(size = 9),   # Adjust axis text size
    legend.position = "none"  # Remove the legend
  )
```


